---
title: "Lab: Synthetic Control Method (SCM)"
subtitle: "Simulated Dataset: California Cap-and-Trade SCM Lab"
author: "EDS241 / ESM244"
format: html
editor: visual
---

# Synthetic Control Method Review

## Case Study: California Cap-and-Trade Policy (2013)

California's cap-and-trade program, launched in 2013, places a cap on total greenhouse gas emissions. Firms must hold permits to emit CO₂, creating a market-based environmental regulation. Our evaluation question is: **Did this policy reduce emissions?**

### The Fundamental Problem

We observe California **with** the cap-and-trade policy. We do **not** observe California **without** the cap-and-trade policy. This is the fundamental problem of causal inference in this setting — we need a **credible counterfactual**.

### Why Not Just Compare California to Other States?

A simple comparison of California to other states runs into several problems: states differ in their energy mix, they experience different economic growth trajectories, they follow different environmental trends, and some states have their own cap-and-trade policies. We need a better counterfactual comparison.

### Core Idea of Synthetic Control

The solution is to create a **"Synthetic California"** — a weighted average of states that did not receive the policy treatment. SCM is a data-driven method that chooses weights so that pre-treatment emissions trends match closely (i.e., pre-2013) and covariates match (income, coal share, electricity price).

### When Do We Use Synthetic Control?

We use SCM when we have **one aggregate unit** that receives the treatment (policy) — for example, a city, state, or nation — and we need a group of comparison units to synthesize the control. In SCM, this comparison group is called the **"donor pool"**.

According to Abadie (2021), SCM works best with:

-   One treated unit
-   Many potential controls
-   A long pre-treatment period (panel data)
-   Not too much pre-treatment volatility
-   Clear intervention timing

### Donor Pool Criteria

Following the approach in Lessmann & Kramer (2024), the donor pool should **exclude** states with cap-and-trade policies during the measurement period or similar energy policies that would affect emissions (i.e., alternative treatment channels). In their study, excluded states included Connecticut, Delaware, Maine, Maryland, Massachusetts, New Hampshire, New Jersey, New York, Rhode Island, and Vermont.

### Seeing the Treatment Effect

If synthetic CA looks like real CA before 2013 (i.e., in emissions), then synthetic CA approximates "what would have happened without cap-and-trade." The treatment effect at each post-treatment period is the gap between observed and synthetic outcomes.

Before the 2013 policy: Real CA $\approx$ Synthetic CA.

After the 2013 policy: the gap equals the treatment effect.

### Pre-Treatment Fit Is Key to Identification

Good pre-treatment fit is the key credibility condition.If we cannot match trends before treatment, we cannot trust the counterfactual after. When evaluating the pre-treatment fit, consider how many measurement time points are available, the volatility of the trend, and the magnitude of deviations between observed and synthetic California.

1.  Why can't we simply pick a single state (e.g., Texas) as a comparison for California? What problems would a simple comparison introduce?

**RESPONSE:** Selecting a single state, like Texas, as a comparison for California because they have different environmental policy attitudes, energy use trends, and economic cycles. These different covariates don't give an apples-to-apples comparison.

2.  What does it mean if the synthetic control closely tracks the treated unit in the pre-treatment period? What if it doesn't?

**RESPONSE:** If the synthetic control matches the treated unit before treatment, it’s a credible counterfactual; if it doesn’t, the post‑treatment results aren’t trustworthy.

# Lab Overview

## Research Question

**Did California's cap-and-trade program (launched in 2013) reduce per capita CO₂ emissions from the power sector?**

-   $Y$ (outcome): Per capita CO₂ emissions from the power sector (`co2_power_pc_tons`)
-   Treated unit: California (`CA`)
-   Treatment year: 2013
-   Donor pool: Other U.S. states (without similar cap-and-trade policies)

This lab uses a simplified example loosely based on the study by Lessmann & Kramer (2024), with simulated data for teaching purposes. The results should not be interpreted literally as simulated data necessarily includes assumptions that omit complexity found in the original study data.

## Predictors

We use the following pre-treatment (2000–2012) averages as predictors to construct the synthetic control:

| Variable            | Description                               | Role      |
|------------------|------------------------------------|------------------|
| `coal_share`        | Share of electricity generation from coal | Predictor |
| `income_pc`         | Per capita income                         | Predictor |
| `elec_price`        | Electricity price                         | Predictor |
| `co2_power_pc_tons` | Pre-treatment average of the outcome      | Predictor |

These variables capture structural differences in energy systems and economies that influence emissions, ensuring the synthetic California is a meaningful comparison.

In this lab, you will:

1.  **Estimate** synthetic California using the `tidysynth` package
2.  **Evaluate pre-treatment fit** by comparing observed vs. synthetic trends
3.  **Conduct placebo tests** to assess whether the estimated effect is statistically meaningful
4.  **Examine the MSPE ratio** for permutation-based inference
5.  **Review the balance table** to check covariate matching

Why do we include the pre-treatment mean of the outcome variable (`co2_power_pc_tons`) as a predictor? What would happen if we only used the other covariates?

**RESPONSE:** Because past emissions contain information about unobserved factors that no covariate can fully capture; without it, the synthetic control won’t match California’s pre‑policy emissions path, and the counterfactual becomes unreliable.

# Analysis

## Setup

#### Load Packages

```{r}
#| warning: false
#| message: false

# install.packages("tidysynth")
# install.packages("tidyverse")
# install.packages("patchwork")

library(tidysynth) # for synthetic control
library(tidyverse)
library(patchwork) # for combining plots
```

## Load Data

The dataset contains annual state-level panel data with CO₂ emissions, energy mix, income, and electricity prices. Each row represents a state-year observation.

```{r}
#| eval: false
#| echo: true

data <- read_csv(here::here("week7", "data", "ca_captrade_scm_sim_data.csv"))

head(data)
```

1.  How many states are in the donor pool? How many years does the data cover?

**RESPONSE:** 49 states, 23 years.

## Declare the Synthetic Control

This is the core step. Using `tidysynth`, we declare the synthetic control specification in a single piped workflow. Let's walk through what each function does:

-   **`synthetic_control()`**: Sets up the SCM framework — specifying the outcome variable, the unit and time identifiers, the treated unit (`"CA"`), the treatment time (`2013`), and requesting placebos for inference.
-   **`generate_predictor()`**: Defines the pre-treatment characteristics used to match the synthetic control to California. We compute 2000–2012 averages of coal share, income, electricity price, and the outcome variable itself.
-   **`generate_weights()`**: Chooses donor weights that minimize the distance between California and the synthetic control on the predictors.
-   **`generate_control()`**: Constructs the synthetic control unit using the optimized weights.

```{r}
scm_data <- data %>%
  synthetic_control(
    outcome = co2_power_pc_tons, # set outcome variable: CO2 per capita
    unit = state, # unit identifier (state)
    time = year, # time identifier (year)
    i_unit = "CA", # intervention (treated) unit: CA
    i_time = 2013, # intervention time: year (2013)
    
    # generate placebo tests for all years
    generate_placebos = TRUE) %>% 
  
  # Define characteristics for synthetic CA to match in order to be considered a good copy of real CA
  generate_predictor(  
    time_window = 2000:2012, # pre treat period to compute predictors
    coal_share = mean(coal_share, na.rm = TRUE), # average share of electricity from coal
    income_pc = mean(income_pc, na.rm = TRUE), #average per capita outcome
    elec_price = mean(elec_price, na.rm = TRUE),# average electricity price
    co2_pre_mean = mean(co2_power_pc_tons, na.rm = TRUE)) %>% # average pre treat outcome
  
  # take the predictor summaries from above and determine weight value to assign each donor state
  generate_weights(
    optimization_window = 2000:2012,# donor weight optimization period
    margin_ipop = 0.02, # margin of precision
    sigf_ipop = 7, # number of signoficant figures
    bound_ipop = 6) %>% 
  
  # Construct synth CA using optimized donor weights
  generate_control() 
```

## Plot Trends

The trends plot is our first visual diagnostic. It overlays the observed outcome for California against synthetic California across the full time series. In the **pre-treatment period** (before the dashed vertical line at 2013), we want to see the two lines tracking closely. In the **post-treatment period**, any divergence represents the estimated treatment effect.

Remember the intuition from lecture: if synthetic CA looks like real CA before 2013, then synthetic CA approximates what would have happened without cap-and-trade.

```{r}
scm_data %>% plot_trends() +
  labs(title = "Observed CA vs Synth CA",
       x = "Year",
       y = "CO2 per capita")


```

1.  Do the observed and synthetic trends track each other well before 2013? What does this tell you about the quality of the synthetic control?

**RESPONSE:** Yes, the observed and synthetic lines track each other very closely in the pre‑2013 period. This indicates **strong pre‑treatment fit**, which is the key credibility condition in synthetic control. When the synthetic line mirrors California before the policy, it means the synthetic control is a **high‑quality counterfactual** for “California without cap‑and‑trade.”

2.  What happens to the gap between the two lines after 2013? What does this suggest about the effect of cap-and-trade?

**RESPONSE:** After 2013, the observed line drops below the synthetic line, and the gap **widens consistently over time**. This pattern suggests that **California’s emissions fell more than they would have without the policy**, implying that cap‑and‑trade likely **reduced per‑capita CO₂ emissions in the power sector**.

## Plot Gaps (Treatment Effect)

The gaps plot shows the **difference** between observed California and synthetic California over time. Before the treatment, the gap should hover near zero (confirming good pre-treatment fit). After 2013, persistent negative gaps would indicate that California's emissions fell below what they would have been without cap-and-trade — i.e., the magnitude of the gap equals the estimated treatment effect.

```{r}
scm_data %>% 
  plot_differences() +
  labs(title = "Gap: Observed - Synthetic")


```

Examine the gap plot. Describe what you observe in both the pre-treatment and post-treatment periods.

**RESPONSE: In the pre‑treatment period (before 2013), The gap hovers very close to zero with only small fluctuations, which means observed and synthetic California were nearly identical before the policy. This indicates excellent pre‑treatment fit and gives you confidence that the synthetic control is a credible counterfactual.

After the intervention line (after 2013), the gap becomes consistently negative and grows larger over time, meaning observed emissions fall below synthetic emissions. This widening negative gap suggests that California’s emissions declined more than they would have without cap‑and‑trade, consistent with a policy‑induced reduction in CO₂ per capita from the power sector.

## Placebo Tests

Now we assess statistical significance using placebo tests. Recall the procedure from lecture: we apply SCM to **every donor state** as if it received the treatment in 2013, then compare California's gap to the distribution of placebo gaps. If California's post-treatment gap stands out clearly from the placebos, this provides evidence that the effect is real and not just noise.

The plot below overlays California's gap (highlighted) against all placebo gaps. Ideally, California's line should diverge noticeably from the placebo distribution after 2013.

```{r}
scm_data %>% 
  plot_placebos() +
  labs(title  = "Placebo Tests") # Synthetic CA and every other state (placebo)
```

Does California's gap stand out from the placebo distribution after 2013?

**RESPONSE:** Yes, California’s gap clearly stands out. In the placebo plot, the California line drops sharply and stays well below zero after 2013, while the placebo states cluster much closer to zero with far smaller movements. That separation shows that California’s post‑treatment gap is unusually large compared to the distribution of placebo gaps, which strengthens the evidence that cap‑and‑trade caused a real reduction in CO₂ emissions rather than the pattern being due to noise or general trends.

## MSPE Ratio Plot

The MSPE ratio provides a more formal way to assess significance. Recall from lecture that the MSPE ratio compares the post-treatment signal to the pre-treatment error:

$$
\text{MSPE Ratio} = \frac{MSPE_{post}}{MSPE_{pre}}
$$

A high ratio means the unit's post-treatment gap is large **relative to its pre-treatment fit**. If California's MSPE ratio ranks at or near the top, the divergence is unlikely due to chance.

```{r}
scm_data %>% 
  plot_mspe_ratio() +
  labs(title = "MSPE Ratio")
```

::: {.callout-tip title="Tip"}
Click the zoom button to open the plot in a new window to better view y axis labels!
:::

1.  Where does California rank in the MSPE ratio distribution? What does this imply about the statistical significance of the estimated effect?

**RESPONSE:** California sits at the very top of the MSPE‑ratio distribution, the largest post‑/pre‑MSPE jump of any state. Because its ratio is so much larger than all donor states, it implies that the divergence between observed and synthetic California after 2013 is far too large to be explained by chance, strengthening the case that the estimated cap‑and‑trade effect is statistically meaningful rather than noise.  

## Balance Table

The balance table compares pre-treatment predictor values for **real California**, **synthetic California**, and the **donor pool average**. Good balance means the synthetic control closely matches California on all predictors — not just the outcome. Large discrepancies between California and its synthetic version would indicate that the optimization did not find a good match, undermining the credibility of the counterfactual.

```{r}
scm_data %>% 
  grab_balance_table()
```

1.  How well does synthetic California match real California on each predictor? Are there any variables where the match is particularly poor?

**RESPONSE:** The only predictor with a noticeable (though still small) mismatch is electricity price, where synthetic CA is a bit higher than actual CA. Everything else is matched almost perfectly, which is exactly what you want for a credible synthetic control.

2.  How does the synthetic control compare to the simple donor pool average?

**RESPONSE:** The donor pool averages are much farther from California than the synthetic control on every predictor. For example: 

Donor average CO₂ pre‑mean: 11.2 (vs CA’s 12.5)
Donor average income: 53,868 (vs CA’s 57,296)
Donor average electricity price: 0.119 (vs CA’s 0.0935)

This shows why SCM is necessary:
A simple average of donor states does not resemble California at all, but the synthetic control does.

# Discussion & Reflection

1.  **Summarize the findings**: Based on the trends plot, gap plot, placebo tests, and MSPE ratio, what is your overall conclusion about the effect of California's cap-and-trade program on per capita CO₂ emissions from the power sector? Does California's emissions fall below synthetic California?

**RESPONSE:** California’s emissions fall clearly below synthetic California after 2013, and across all diagnostics the evidence strongly supports that cap‑and‑trade caused a real reduction in per‑capita CO₂ emissions from the power sector.

2.  **Diagnostics checklist**: Review the four criteria from lecture — strong pre-treatment fit, transparent donor pool, no spillover, and favorable placebo comparison. How does our analysis perform on each?

**RESPONSE:** We have excellent pre‑treatment fit, a transparent donor pool, no evidence of spillover, and placebo tests that strongly favor California’s effect as unusually large.

3.  **Donor pool selection**: Why is it important that donor states did not implement similar carbon pricing policies during the study period? What would happen to our estimates if some donor states also adopted cap-and-trade (hint: think about spillover and the no-interference assumption)?

**RESPONSE:** If donor states also adopted carbon pricing, their emissions would drop too, contaminating the counterfactual and violating the no‑interference assumption, which would bias the estimated effect toward zero.

4.  **Comparison to other methods**: How is SCM different from matching? What are the advantages of building a weighted synthetic counterfactual from the donor pool to match the treated unit's pre-treatment outcome path, versus matching on observed covariates?

**RESPONSE:** SCM differs from matching because it builds a weighted combination of states to reproduce the treated unit’s entire pre‑treatment outcome path, giving a far more credible counterfactual than matching only on observed covariates.
